\documentclass[a4paper]{article}

\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{hyperref}

\usepackage{listings}
\lstset{%
    language=Java,
}

\hypersetup{%
    pdfauthor = {Martin Barksten, Victor Koronen},
    pdftitle = {Compiler Construction},
    pdfsubject = {DD2488},
    pdfkeywords = {compiler},
    pdfcreator = {LaTeX with hyperref package},
    pdfproducer = {pdflatex}
}

\title{Compiler Construction}
\author{%
    Martin Barksten <\href{mailto:barksten@kth.se}{barksten@kth.se}> \and
    Victor Koronen <\href{mailto:koronen@kth.se}{koronen@kth.se}>
}
\date{\today}

\begin{document}

\maketitle
\thispagestyle{empty}

\section{Introduction}

This document describes the design and implementation of a compiler for a subset
of the Java language know as MiniJava \cite{minijavaproject} (with some
adaptations \cite{dd2488project} for the course DD2488 at KTH).

The document begins with a section about the structure of the program, followed
by a section giving an overview of the code and finally a section about how to
build, test and run the compiler. Included in the appendix is a section about some
of the trickier bugs encountered.

\section{Structure}

This section gives a description of how the program is structured. The sections
come in the order of steps when compiling, beginning with the lexing and parsing
steps and ending with the assembling. The compiler compiles to the JVM with help
from the Jasmin assembler \cite{jasmin}. Compiling to the JVM with Jasmin
allows us to skips a lot of common backend steps, such as Immediate Representation,
Control Flow Analysis and Register Allocation.

\subsection{Lexer and Parser}

Lexing and parsing is done using JavaCC \cite{javacc}, a Compiler Compiler for
Java that generates a LL-parser. We also considered using ANTLR \cite{antlr}, a
parser generator that generates LR-parsers. After some experimentation with both
libraries we decided to use JavaCC, mainly because we got the impression that
JavaCC is more well-used and better documented.

JavaCC handles tokenization and allows the programmer to write an LL grammar in
\texttt{.jj} files, which JavaCC then ``compile'' to Java files. In order
to be able to use a JavaCC-generated LL-parser the given grammar had to
have all left-recursion eliminated, and the \textit{Exp} operators had to be given the
right precedence.

Eliminating left-recursion can be done in JavaCC by using the
ability to match $0$ or more possible productions. The grammar

$$A \rightarrow A + B$$

can be converted to

$$A \rightarrow B (+ B)*$$

where star means it should match $0$ or more times. When rewriting it this way
it is clear that the left recursion has been eliminated.

The other problem was to give operators the right precedence. Assume that we have the
following grammar

$$A \rightarrow B + B$$
$$A \rightarrow B \times B$$

and we want $\times$ to have higher precedence than $+$. We can simply rewrite it as

$$A \rightarrow B (+ B)*$$
$$B \rightarrow C (\times C)*$$

which will give $\times$ higher precedence than $+$.

\subsection{Abstract Syntax Tree}

Constructing the Abstract Syntax Tree (AST) first requires creating all classes
necessary to represent it, one class for each unique terminator. The AST is
constructed during the parsing step using JavaCC. When, for example, a \texttt{Plus}
operator is found, we will continue the parsing to find the left and right
expression and then create an instance of the \texttt{Plus} class. This object is then
returned from the grammar rule.

\subsection{Type checking}

The type checking is done in two passes. The first pass binds all methods and
variables to their types, and the second pass checks that all statements and
expressions operate on the right types. In order to perform these two steps we
need to recursively visit the entire AST, which we do by using the visitor
pattern.

The visitor pattern allows us to keep all the traversal logic in a single class,
as opposed to having it spread out across all AST node classes
\cite{oodesignvisitor}.

In order to save the types we use a symbol table to map an identifier to a
corresponding type. We implement the symbol table using three different
tables: a program table, a class table and a method table. Each table saves
slightly different information, which is why we chose to implement it in this
fashion. The program table maps all class identifiers to their respective class
table. The class tables maps the method identifiers to their respective method
tables and also keeps track of what fields the class has, as well as an optional
super class. Finally the method table maps variable identifiers to their types.
The method table also keeps track of which variables are locals and which are
parameters.

When binding identifiers to types we traverse the tree and whenever we encounter
the declaration of a class, method, field or variable we add an entry to the
proper table. Classes are added to the program table, methods and fields to the
corresponding class table and variables to the corresponding method table.

The program table, allowing access to all other tables, is then passed to the
visitor used for the second pass. In the second pass we once again traverse the
tree, checking that all operations on variables is correct. Additionally when
calling a method we check that the object it is called on has the method defined
for it. This is once again done using a visitor, and traversing the tree
recursively.

\subsection{Assembling}

The assembly step is done using Jasmin \cite{jasmin}. The reason is that
Jasmin is, as the authors themselves say ``the de-facto standard assembly format for Java''.

Assembling is done by recursively traversing the AST, translating each node into Jasmin
bytecode. The translation is done by first visiting all child nodes and getting their
bytecode, then the node uses their code to generate its own byte. An example of
this is the following code

\begin{lstlisting}
if(true) {
    a = 1;
} else {
    a = 2;
}
\end{lstlisting}

Which has an AST roughly corresponding to \texttt{if(true, (a = 1), (a = 2))}.
When assembling this node it will first get the bytecode for \texttt{true}, \texttt{a = 1}
and \texttt{a = 2}. It will then use these to create the proper if statement, which is
to load \texttt{true} and then do a branch instruction executing either \texttt{a = 1} or
\texttt{a = 2}.

In order to keep the assembling simple, all \texttt{Exp}s will leave the value behind at the head
of the stack, as could be expected.

Two things to keep in mind during the assembling to bytecode is labels and the stack size.

Labels need to be unique in order to not encounter strange and undefined behavior. This is done
by saving how many labels have been created with a specific name and when
a node requires a new label the name given has a unique number appended to ensure uniqueness.

The other problem -- calculating the stack size -- is considerably harder. The first problem
is to calculate for each operation how large a stack it will require at most. This is
rather straightforward, as an example a method call will load the object being called
and all parameters for the method. This will require a stack size of $p + 1$ where $p$
is the number of parameters for the method.

The tricky part is that the stack may grow during the calculation of values. As an example
assume we have the following code:

\begin{lstlisting}
obj.method(3 * 3)
\end{lstlisting}

If we just calculate the stack size of each separate operation we will find we need only
a stack of size $1 + 1 = 2$ according to the formula given above. However, if we actually
check we will see that first \texttt{obj} will be pushed and then the bytecode for $3 * 3$,
which will push \texttt{3} twice before doing an operation on them. This means we need a
stack of size $1+2 = 3$.

Thus in order to correctly calculate the stack size the compiler must keep track of how
large the stack currently is. Then each node calculates its required stack size, which is
the current stack size plus the number of values pushed by the node. The maximum required
stack size by all nodes inside a method is the methods required maximum stack size.

\section{Code overview}

\begin{description}
\item[\texttt{frame}] Package containing interfaces for abstractions of frames,
    records and variable accesses.
    \begin{description}
        \item[\texttt{frame.VMAccess}] Interface for declaring, loading and
            storing a variable on a VM.
        \item[\texttt{frame.VMFrame}] Interface for handling frames on a VM.
        \item[\texttt{frame.VMRecord}] Interface for handling heap objects on a
            VM.
    \end{description}
\item[\texttt{jvm}] Package containing implementations of the frame, record and
    variable access interfaces for the JVM.
    \begin{description}
        \item[\texttt{jvm.Frame}] Implementation of the frame handling interface
            for the JVM.
        \item[\texttt{jvm.Hardware}] Implementations of variable and method
            signatures for the JVM.
        \item[\texttt{jvm.Record}] Implementation of the heap object interface
            for the JVM.
    \end{description}
\item[\texttt{mjc}] Package containing the executable main class and the parser
    code.
    \begin{description}
        \item[\texttt{mjc.JVMMain}] The main JVM executable. Parses command line
            parameters, instantiates collaborator objects (such as the parser)
            and orchestrates interaction between the collaborators to produce
            the desired result.
        \item[\texttt{mjc.Parser}] The JavaCC-generated LL-parser.
    \end{description}
\item[\texttt{symbol}] Package containing abstractions of symbols and the symbol
    table.
    \begin{description}
        \item[\texttt{symbol.Symbol}] The abstraction of a symbol.
        \item[\texttt{symbol.ProgramTable}] The program symbol subtable.
        \item[\texttt{symbol.ClassTable}] The class symbol subtable.
        \item[\texttt{symbol.MethodTable}] The method symbol subtable.
    \end{description}
\item[\texttt{syntaxtree}] Package containing class definitions for all AST
    nodes.
\item[\texttt{temp}] Package containing abstractions of assembly level labels
    and variables.
    \begin{description}
        \item[\texttt{temp.Label}] The abstraction of a label.
        \item[\texttt{temp.Temp}] The abstraction of a variable.
    \end{description}
\item[\texttt{visitor}] Package containing implementations of visitors used to
    traverse tree structures, such as the AST.
    \begin{description}
        \item[\texttt{visitor.TypeDefVisitor}] Performs the first part of type
            checking. Traverses the AST and builds a symbol table structure.
        \item[\texttt{visitor.TypeCheckVisitor}] Performs the second part of
            type checking. Traverses the AST and checks types with help from the
            aforementioned symbol table.
        \item[\texttt{visitor.JVMVisitor}] Traverses the AST and writes Jasmin
            assembly.
    \end{description}
\end{description}

\section{Building, testing and running}

The compiler can be built using the Java build tool Apache Ant\texttrademark.
Just issue the command \texttt{ant} from the compiler root directory and Ant
will generate the parser using JavaCC, compile the compiler and compress all JVM
bytecode into a JAR archive.

Testing the compiler can be done by using the script \texttt{./script/test}. It
will first invoke Ant to ensure the compiler has been built and then proceed to
test the compiler against the test cases found in the directory
\texttt{testcases/}.

Running the compiler can be done by using the script \texttt{./script/run}.
Given a MiniJava source file it will try to compile the file using our compiler
and then to execute the generated program using the system Java runtime.

\appendix

\section{Parsing bugs}
In this section we describe the bugs that caused the most problems during the creation of the program.

\subsection{Binding of \textit{Not}}

During parsing the most interesting bug was the binding of \textit{Not}. As an
example, consider the following piece of code.

\begin{lstlisting}
!false && true
\end{lstlisting}

This expression above would incorrectly be parsed as

\begin{lstlisting}
!(false && true)
\end{lstlisting}

This caused some strange behavior that turned out to be very difficult to debug.
This bug was not found until work began on the JVM assembly stage. Finding what
caused the bug caused considerable headache both because we assumed our parser
was correct and because almost all boolean logic was incorrect (more about that in
the section about JVM assembly bugs). Thus two bugs became entangled which made
it hard to find either of the two.

Once we discovered that the bug was in the parser resolving it was easy.

The thing we learned from this was to debug more in earlier steps, and if we don't
be less trusting of them. Unnecessary time was spent debugging the JVM assembler
because we didn't test the parser enough.

\section{Branching logic in the JVM}

When implementing the JVM assembly step, several bugs related to logical
operators were found. The bug would arise when having code as below (note
the use of the \texttt{Not} operator here).

\begin{lstlisting}
!(1 < 2)
\end{lstlisting}

The expression would incorrectly evaluate as false. Finding why the
bug arose was easy -- branching on true or false was clearly wrong. However,
fixing it took quite some time due to how the branching works.

The JVM assembly instruction \texttt{ifeq} compares to zero, which we internally
used as false. So if \texttt{ifeq} branched, the value was false rather than
true. This behavior made every single node in the AST related to booleans behave
incorrectly.

Other minor bugs also arose when converting to bytecode, however these were all
fixed faster and easier.

\begin{thebibliography}{99}

\bibitem{minijavaproject}
    J.~Cangussu, J.~Palsberg, V.~Samanta,
    \emph{Modern Compiler Implementation in Java: the MiniJava Project},
    <\url{http://www.cambridge.org/resources/052182060X/}>, retrieved 2014-04-27

\bibitem{dd2488project}
    T.~Granlund,
    \emph{Project in DD2488, Compiler Construction (komp14)},
    <\url{http://www.csc.kth.se/utbildning/kth/kurser/DD2488/komp14/project/}>, retrieved 2014-04-27

\bibitem{jasmin}
    J.~Meyer,
    \emph{JASMIN HOME PAGE} (2004),
    <\url{http://jasmin.sourceforge.net/}>, retrieved 2014-04-27

\bibitem{javacc}
    JavaCC contributors,
    \emph{Java Compiler Compiler\texttrademark (JavaCC\texttrademark) -- The Java Parser Generator},
    <\url{https://javacc.java.net/}>, retrieved 2014-04-27

\bibitem{antlr}
    ANTLR~/~T.~Parr,
    \emph{ANTLR},
    <\url{http://www.antlr.org/}>, retrieved 2014-05-08

\bibitem{oodesignvisitor}
    A.~Ianculescu,
    \emph{Visitor Pattern},
    <\url{http://www.oodesign.com/visitor-pattern.html}>, retrieved 2014-05-08

\end{thebibliography}

\end{document}
